% Abstract to go at the start of my report. 
% Word Limit: 500 words. 
% May also have some ethics statement, etc... 
% Whatever that is supposed to mean...

Real-time systems are an area of increasing relevance and interest in the field 
of Computer Science. Generally stated, a real-time system is a type of information 
processing system in which a response to an externally generated input stimuli 
has to be generated within a finite and specified time period. Correctness of the 
system may depend not only the result, but also the time that it was delivered. 
Such systems are commonplace and are found in a wide range of 
situations such as manufacturing processes, automotive vehicles and embedded 
sensoring.

The development of real-time systems typically involves the use of a programming 
language providing explicit support for real-time systems. 
This support involves the provision of protocols that allow the system to 
behave in a more reliable manner, such as asynchronous transfer of control or 
bounded priority inversion. 
My project entails the provision of such support to a new programming language: 
D. D currently has very limited support for the development of real-time systems. 

Several programming languages already exist that provide support for the 
development of real-time systems. The most notable of which are 
real-time Java, Ada and C with POSIX Threads. These languages are well researched, 
with a rich field of study comparing their approaches and behaviour. 
However, comparatively little research has gone into providing new programming 
languages with support for the development of real-time systems. My project 
targets this omission by extending real-time support a previously unsupported 
language.

By examining existing programming languages and research on real-time systems, a core set 
of functionality for any real-time system supporting language was defined. 
In some aspects, such as concurrent programming, support was already provided in D. 
However, for the majority of real-time programming language features,
additional support was required. Different approaches for the implementation 
of these features was considered, and functionality was implemented into 
a library. With regards to asynchronous transfer of control, there was no clear 
approach for implementation. As a result, two alternative approaches have 
been considered, profiled and implemented. A comparison between the two approaches 
has shown that a POSIX signal based approach leads to a higher performance 
solution. 

Furthermore, a series of experiments compares behaviour of a D program written 
using the added features with an equivalent real-time Ada program. This found 
that\ldots. 

As a result of this project, the pool of applicable languages for the development 
of real-time systems has been extended, incorporating a new and powerful language, D, which 
boasts high performance, coupled with strong safety. 


%However, there is no reason that this pool of applicable languages can not be extended to 
%incorporate new, powerful programming languages. 
%D is one such language, aiming for a C++ style syntax, coupled with powerful 
%abstractions to ease development and the ability to take advantage of modern 
%compiler technology to provide aggressively optimised code.
%The D programming language is of particular interest for 
%the development of real-time applications due to it's combined ease of development
%and rich feature set. 

%However, it is noteworthy that D does not provide support for real-time systems: 
%it has the required characteristics of enabling concurrent programming and low-level 
%control, but lacks required features. 
%In order to develop a real-time application or system, additional functionality 
%must therefore be provided.
%This project intends to contribute to the pool of languages supporting the development 
%of real-time systems and applications by providing necessary features 
%to the D Programming language. 


